---
title: "Homework1"
author: "David Lawlor"
date: "2026-06-09"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# connect R studio and github

Did most of the setup in hw0

## 12.2

hw0.

## 12.3

Just put the url into rstudio worked fine

## 12.4

wen't fine forgot to switch credentials easily fixed

## 12.5

Pushed fine no problems

## 12.6

Checked and saw my changes on the repo on github

## Question 1

install.packages("mclust") install.packages("ggplot2")

```{r}

library(mclust) 
library(ggplot2) 
data(banknote)

str(banknote)
```

## Question 2

```{r}

ggplot(banknote, aes(x = Status, fill = Status)) + geom_bar()
```

This is a barchart with Status on the x-axis and fill color. There are
exactly 100 genuine and 100 counterfeit banknotes. No anomalies.

## Question 3

```{r}

ggplot(banknote, aes(x = Diagonal, fill = Status)) +
geom_histogram(binwidth = 0.5, position = "identity", alpha = 0.7)
```

This is a histogram with Diagonal on the x-axis and Status as fill
color. Genuine banknotes have a bigger diagonal than counterfeit ones.
There is a small overlap in the middle where the two groups mix.

## 4

```{r}

ggplot(banknote, aes(x = Right, y = Diagonal, color = Status)) +
geom_point()
```

This is a scatterplot with Right on the x-axis, Diagonal on the y-axis,
and status as the color. The two groups are separate clusters, meaning
these two measurements alone can tell apart genuine and counterfeit
banknotes. There are a few outliers on both sides, but the two groups
still separate.
