---
title: "What Team Statistics Are Most Associated with Winning NBA Games?"
author: "David Lawlor"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.path = "figures/")
```

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(knitr)
```

# Introduction

Winning an NBA game comes down to a lot of things, but a box score only
tracks a handful of them: shooting, rebounds, assists, turnovers, steals,
blocks, and fouls. We wanted to know which of these are actually tied to
winning the most.

Questions we're looking to answer:

1. Which team statistics have the strongest relationship with winning?
2. Are rebounds, assists, or shooting percentages more important for success?
3. Do winning teams commit fewer turnovers?
4. Which statistics differ the most between winning and losing teams?

## Team Members

- David Lawlor

# Data

## Data Source

We used team level box scores from NBA regular season games, 2010-11 through
2023-24, pulled from a public dataset on GitHub
([`NocturneBear/NBA-Data-2010-2024`](https://github.com/NocturneBear/NBA-Data-2010-2024),
MIT licensed). Each game has two rows in the data, one for each team.

We saved the data as a CSV in `data/` instead of pulling it live with a
package like `hoopR`, so this report doesn't depend on an API being up when
it knits.

```{r load-data}
nba_raw <- read_csv("data/regular_season_totals_2010_2024.csv",
                     show_col_types = FALSE)
dim(nba_raw)
```

## Data Cleaning

The raw file has a lot of columns we don't need (rank columns for every
single stat). We kept the main identifying columns and box score stats, made
a numeric win column out of `WL`, and made a home/away column by checking if
`MATCHUP` has an `"@"` in it (away games look like `"MIL @ MIA"`).

```{r clean-data}
nba <- nba_raw %>%
  transmute(
    season    = SEASON_YEAR,
    team      = TEAM_NAME,
    game_id   = GAME_ID,
    game_date = as.Date(GAME_DATE),
    matchup   = MATCHUP,
    home      = !str_detect(MATCHUP, "@"),
    win       = as.integer(WL == "W"),
    fg_pct    = FG_PCT,
    fg3_pct   = FG3_PCT,
    ft_pct    = FT_PCT,
    oreb = OREB, dreb = DREB, reb = REB,
    ast = AST, tov = TOV, stl = STL, blk = BLK,
    pf  = PF,  pts = PTS
  ) %>%
  filter(!is.na(win))

# Any missing values left after cleaning?
colSums(is.na(nba))
```

No missing values left in the columns we use. One thing worth noting: a few
teams changed cities or names during this stretch (the Charlotte Bobcats
became the Hornets in 2014, for example), so `team` isn't a perfectly
consistent identifier across all 14 seasons. That doesn't affect this
analysis since we never group by team, but it would matter for a team-level
question.

## Marginal Summaries

```{r summaries}
glimpse(nba)

n_distinct(nba$game_id)
n_distinct(nba$season)
n_distinct(nba$team)

summary(nba %>% select(fg_pct, fg3_pct, reb, ast, tov, stl, blk, pts))
```

This is `r n_distinct(nba$season)` seasons,
`r format(n_distinct(nba$game_id), big.mark = ",")` games, and
`r nrow(nba)` team-game rows, split about evenly between wins and losses
since every game has one winner and one loser.

## Points Distribution

```{r pts-hist}
ggplot(nba, aes(pts)) +
  geom_histogram(binwidth = 5, fill = "#1d428a", color = "white") +
  labs(
    title = "Distribution of Team Points per Game (2010-11 to 2023-24)",
    x = "Points Scored", y = "Number of Team-Games"
  ) +
  theme_minimal()
```

Points per game are roughly bell shaped, centered a bit above 100, with a
few high scoring games pulling out a right tail.

# Analysis

## Home Court Advantage

Before getting into box score stats, we checked something simpler: does
playing at home matter?

```{r home-advantage}
home_summary <- nba %>%
  group_by(home) %>%
  summarise(win_rate = mean(win), games = n(), .groups = "drop") %>%
  mutate(location = ifelse(home, "Home", "Away"))

kable(home_summary %>% select(location, win_rate, games),
      digits = 3, col.names = c("Location", "Win Rate", "Games"),
      caption = "Win rate by home/away status")
```

```{r home-advantage-chart}
ggplot(home_summary, aes(x = location, y = win_rate, fill = location)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Away" = "#c8102e", "Home" = "#1d428a")) +
  labs(title = "Home Teams Win More Often",
       x = NULL, y = "Win Rate") +
  theme_minimal() +
  theme(legend.position = "none")
```

Home teams won about
`r scales::percent(home_summary$win_rate[home_summary$location == "Home"], accuracy = 0.1)`
of games, compared to
`r scales::percent(home_summary$win_rate[home_summary$location == "Away"], accuracy = 0.1)`
for away teams. That's not really a "stat" a team controls in the moment, so
we added home/away as a control in the regression later on.

## Win vs. Loss Averages

```{r win-loss-table}
summary_tbl <- nba %>%
  mutate(result = ifelse(win == 1, "Win", "Loss")) %>%
  group_by(result) %>%
  summarise(
    `FG%` = mean(fg_pct), `3P%` = mean(fg3_pct), `FT%` = mean(ft_pct),
    Reb = mean(reb), Ast = mean(ast), TOV = mean(tov),
    Stl = mean(stl), Blk = mean(blk), PF = mean(pf),
    .groups = "drop"
  )

kable(summary_tbl, digits = 3, caption = "Average box score stats: wins vs. losses")
```

Winning teams shoot better, grab more rebounds, get more assists and
steals, and turn the ball over less. None of that is shocking. The real
question is how much each one matters, and whether some of these numbers
are just tied to each other.

```{r fg-pct-boxplot}
ggplot(nba, aes(x = factor(win, labels = c("Loss", "Win")), y = fg_pct,
                 fill = factor(win, labels = c("Loss", "Win")))) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("Loss" = "#c8102e", "Win" = "#1d428a")) +
  labs(title = "Field Goal Percentage: Wins vs. Losses",
       x = NULL, y = "Field Goal %") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Which Stats Correlate With Winning

```{r marginal-cor}
vars <- c("fg_pct", "fg3_pct", "ft_pct", "oreb", "dreb", "reb",
          "ast", "tov", "stl", "blk", "pf")

marginal_cor <- tibble(
  stat = vars,
  correlation = sapply(vars, function(v) cor(nba[[v]], nba$win))
) %>%
  arrange(desc(correlation))

kable(marginal_cor, digits = 3,
      caption = "Correlation of each stat with winning, one at a time")
```

```{r marginal-cor-chart, fig.height=5}
ggplot(marginal_cor, aes(x = reorder(stat, correlation), y = correlation,
                          fill = correlation > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#1d428a", "FALSE" = "#c8102e"), guide = "none") +
  labs(title = "Marginal Correlation of Each Box Score Stat With Winning",
       x = NULL, y = "Correlation with Win") +
  theme_minimal()
```

Field goal %, defensive rebounds, three point %, and assists have the
strongest positive correlation with winning. Turnovers and fouls are the
only negative ones. Offensive rebounds barely correlate at all, which is a
little surprising since extra possessions should help.

## Are These Stats Actually Independent?

These stats aren't independent of each other. An assist can only happen
after a made shot, so assists and field goal % should move together for
reasons that have nothing to do with winning specifically. To check this
properly, we ran a logistic regression predicting wins from all the stats
at once, which controls for that overlap instead of looking at each stat
alone.

```{r multicollinearity}
predictors <- c("fg_pct", "fg3_pct", "oreb", "dreb", "ast", "tov", "stl", "blk", "pf")
cor_matrix <- cor(nba[, predictors])
diag(cor_matrix) <- NA

cat("Strongest correlation between two predictors:",
    round(max(abs(cor_matrix), na.rm = TRUE), 3),
    "(assists and field goal %)\n")
```

```{r logistic-regression}
nba_scaled <- nba %>%
  mutate(across(all_of(predictors), ~ as.numeric(scale(.)))) %>%
  mutate(home_num = as.integer(home))

fit <- glm(
  win ~ fg_pct + fg3_pct + oreb + dreb + ast + tov + stl + blk + pf + home_num,
  data = nba_scaled, family = binomial()
)

summary(fit)

mcfadden_r2 <- 1 - fit$deviance / fit$null.deviance
cat("McFadden pseudo R-squared:", round(mcfadden_r2, 3), "\n")
```

```{r regression-chart, fig.height=5}
coef_tbl <- tibble(
  stat = names(coef(fit))[-1],
  estimate = coef(fit)[-1]
) %>%
  arrange(desc(estimate))

ggplot(coef_tbl, aes(x = reorder(stat, estimate), y = estimate,
                      fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#1d428a", "FALSE" = "#c8102e"), guide = "none") +
  labs(title = "Standardized Logistic Regression Coefficients",
       subtitle = "Association with winning, holding other stats constant",
       x = NULL, y = "Coefficient (log-odds per 1 SD)") +
  theme_minimal()
```

Once we control for everything else, assists flips from a strong positive
correlation to a small negative coefficient, since its positive effect was
mostly just riding along with field goal %. FG% and defensive rebounds stay
the strongest positive predictors, and turnovers stay negative. The model's
McFadden pseudo R-squared is about `r round(mcfadden_r2, 2)`, so these stats
explain a good chunk of who wins, but not all of it. Also worth keeping in
mind: this shows what's associated with winning, not what happened before
tip-off or what causes what.

# Conclusions

Going back to our original questions:

1. **Which team statistics have the strongest relationship with winning?**
   Field goal % and defensive rebounds, both alone and after controlling
   for everything else. Turnovers and fouls are the only stats that are
   consistently negative.

2. **Are rebounds, assists, or shooting percentages more important?**
   Shooting percentage matters most, along with defensive rebounding.
   Assists look important alone, but most of that effect turns out to
   just be shooting percentage in disguise.

3. **Do winning teams commit fewer turnovers?**
   Yes. Winning teams average
   `r round(summary_tbl$TOV[summary_tbl$result == "Win"], 1)`
   turnovers a game compared to
   `r round(summary_tbl$TOV[summary_tbl$result == "Loss"], 1)`
   for losing teams, and turnovers stay negative even after controlling
   for every other stat.

4. **Which statistics differ the most between winning and losing teams?**
   Field goal % and defensive rebounds show the biggest, most consistent
   gap. Offensive rebounds were the biggest surprise: no real correlation
   alone, but a real positive effect once shooting percentage is held
   constant, probably because offensive rebounds are more available after
   a missed shot in the first place.

Overall this backs up a pretty intuitive story: shoot well, control the
defensive glass, take care of the ball. But we only got there by checking
that these correlations weren't just riding on each other.

# Reproducibility

This report reads `data/regular_season_totals_2010_2024.csv` straight from
this repo, so it knits without needing internet access. It only uses
tidyverse packages from class (`dplyr`, `tidyr`, `readr`, `stringr`,
`ggplot2`, `knitr`).

```{r session-info}
sessionInfo()
```
