---
title: "What Team Statistics Are Most Associated with Winning NBA Games?"
author: "David Lawlor"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.path = "figures/")
```

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(knitr)
```

# Introduction

Winning in the NBA is the product of dozens of small factors on both ends of
the floor, but box scores only record a handful of them: shooting
percentages, rebounds, assists, turnovers, steals, blocks, and fouls. This
project asks a simple question with a not-so-simple answer: **of the
statistics recorded in a standard NBA box score, which ones are most closely
associated with a team winning the game?**

Specifically, we investigate:

1. Which team statistics have the strongest relationship with winning?
2. Are rebounds, assists, or shooting percentages more important for success?
3. Do winning teams commit fewer turnovers?
4. Which statistics differ the most between winning and losing teams?

## Team Members

- David Lawlor

# Data

## Data Source

The analysis uses team-level game totals for NBA regular-season games from
the 2010-11 through 2023-24 seasons, originally sourced from the NBA Stats
API and distributed as a static CSV in the public
[`NocturneBear/NBA-Data-2010-2024`](https://github.com/NocturneBear/NBA-Data-2010-2024)
dataset (MIT licensed). Each row is one team's full box score line for one
game (so every game contributes two rows: one for each team).

We use this static, versioned CSV rather than a live-scraping package (e.g.
`hoopR`) so that this document knits the same way every time without
depending on an external API being reachable at render time — the CSV is
committed alongside this report in `data/`.

```{r load-data}
nba_raw <- read_csv("data/regular_season_totals_2010_2024.csv",
                     show_col_types = FALSE)
dim(nba_raw)
```

## Data Cleaning

The raw file includes a lot of columns we don't need (league-wide rank
columns for every stat). We keep the identifying columns and the core box
score counting stats, derive a numeric win indicator from `WL`, and derive a
home/away indicator by checking whether the `MATCHUP` string contains `"@"`
(away games are formatted like `"MIL @ MIA"`; home games like `"GSW vs.
POR"`).

```{r clean-data}
nba <- nba_raw %>%
  transmute(
    season    = SEASON_YEAR,
    team      = TEAM_NAME,
    game_id   = GAME_ID,
    game_date = as.Date(GAME_DATE),
    matchup   = MATCHUP,
    home      = !str_detect(MATCHUP, "@"),
    win       = as.integer(WL == "W"),
    fg_pct    = FG_PCT,
    fg3_pct   = FG3_PCT,
    ft_pct    = FT_PCT,
    oreb = OREB, dreb = DREB, reb = REB,
    ast = AST, tov = TOV, stl = STL, blk = BLK,
    pf  = PF,  pts = PTS
  ) %>%
  filter(!is.na(win))

# Any missing values left after cleaning?
colSums(is.na(nba))
```

No missing values remain in the columns we use. One known limitation: a few
franchises changed names or cities during this period (e.g. the Charlotte
Bobcats became the Hornets in 2014), so `team` is not a perfectly stable
identifier across all 14 seasons. That doesn't affect this analysis, since we
never aggregate by team identity, but it would matter for any team-level
(rather than game-level) question.

## Marginal Summaries

```{r summaries}
glimpse(nba)

n_distinct(nba$game_id)
n_distinct(nba$season)
n_distinct(nba$team)

summary(nba %>% select(fg_pct, fg3_pct, reb, ast, tov, stl, blk, pts))
```

The dataset covers `r n_distinct(nba$season)` seasons,
`r format(n_distinct(nba$game_id), big.mark = ",")` unique games, and
`r nrow(nba)` team-game rows (roughly balanced between wins and losses, since
every game has exactly one winner and one loser).

## Example Visualization

```{r pts-hist}
ggplot(nba, aes(pts)) +
  geom_histogram(binwidth = 5, fill = "#1d428a", color = "white") +
  labs(
    title = "Distribution of Team Points per Game (2010-11 to 2023-24)",
    x = "Points Scored", y = "Number of Team-Games"
  ) +
  theme_minimal()
```

Team scoring is roughly bell-shaped and centered a little above 100 points,
with a long-ish right tail from occasional blowout-pace games.

# Analysis

## Home Court Advantage

Before looking at box-score stats, it's worth checking a simpler, well-known
effect: does playing at home matter?

```{r home-advantage}
home_summary <- nba %>%
  group_by(home) %>%
  summarise(win_rate = mean(win), games = n(), .groups = "drop") %>%
  mutate(location = ifelse(home, "Home", "Away"))

kable(home_summary %>% select(location, win_rate, games),
      digits = 3, col.names = c("Location", "Win Rate", "Games"),
      caption = "Win rate by home/away status")
```

```{r home-advantage-chart}
ggplot(home_summary, aes(x = location, y = win_rate, fill = location)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Away" = "#c8102e", "Home" = "#1d428a")) +
  labs(title = "Home Teams Win More Often",
       x = NULL, y = "Win Rate") +
  theme_minimal() +
  theme(legend.position = "none")
```

Home teams won about
`r scales::percent(home_summary$win_rate[home_summary$location == "Home"], accuracy = 0.1)`
of games versus
`r scales::percent(home_summary$win_rate[home_summary$location == "Away"], accuracy = 0.1)`
for away teams. This is a real and well-documented effect, and it's a useful
reminder before we get to box-score stats: some of what predicts winning
isn't a "stat" a team controls in the moment at all. We include `home` as a
control variable in the regression below for that reason.

## How Do Winning and Losing Teams Differ, On Average?

```{r win-loss-table}
summary_tbl <- nba %>%
  mutate(result = ifelse(win == 1, "Win", "Loss")) %>%
  group_by(result) %>%
  summarise(
    `FG%` = mean(fg_pct), `3P%` = mean(fg3_pct), `FT%` = mean(ft_pct),
    Reb = mean(reb), Ast = mean(ast), TOV = mean(tov),
    Stl = mean(stl), Blk = mean(blk), PF = mean(pf),
    .groups = "drop"
  )

kable(summary_tbl, digits = 3, caption = "Average box score stats: wins vs. losses")
```

Winning teams out-shoot losing teams from the field and from three, grab
more rebounds, record more assists and steals, and commit fewer turnovers
and fouls. None of that is surprising on its own — the question is *how
much* each of these differs, and whether some of these differences are just
echoes of the others (e.g. assists naturally go up when shooting is better,
since a made basket is a prerequisite for an assist).

```{r fg-pct-boxplot}
ggplot(nba, aes(x = factor(win, labels = c("Loss", "Win")), y = fg_pct,
                 fill = factor(win, labels = c("Loss", "Win")))) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("Loss" = "#c8102e", "Win" = "#1d428a")) +
  labs(title = "Field Goal Percentage: Wins vs. Losses",
       x = NULL, y = "Field Goal %") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Which Stats Correlate Most With Winning, On Their Own?

```{r marginal-cor}
vars <- c("fg_pct", "fg3_pct", "ft_pct", "oreb", "dreb", "reb",
          "ast", "tov", "stl", "blk", "pf")

marginal_cor <- tibble(
  stat = vars,
  correlation = sapply(vars, function(v) cor(nba[[v]], nba$win))
) %>%
  arrange(desc(correlation))

kable(marginal_cor, digits = 3,
      caption = "Marginal (one-at-a-time) correlation of each stat with winning")
```

```{r marginal-cor-chart, fig.height=5}
ggplot(marginal_cor, aes(x = reorder(stat, correlation), y = correlation,
                          fill = correlation > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#1d428a", "FALSE" = "#c8102e"), guide = "none") +
  labs(title = "Marginal Correlation of Each Box Score Stat With Winning",
       x = NULL, y = "Correlation with Win") +
  theme_minimal()
```

On their own, field goal percentage, defensive rebounds, three-point
percentage, and assists show the strongest positive association with
winning; turnovers and personal fouls are the only stats negatively
associated with winning. Offensive rebounds are essentially uncorrelated
with winning at the marginal level — which seems surprising at first (more
extra possessions should help, shouldn't it?).

## Being Skeptical: Are These Stats Really Independent?

A one-at-a-time correlation can be misleading, because these stats are not
independent of each other. For example, an assist can only happen after a
made shot, so we'd expect assists and field goal percentage to move
together for reasons that have nothing to do with winning specifically.

```{r multicollinearity}
predictors <- c("fg_pct", "fg3_pct", "oreb", "dreb", "ast", "tov", "stl", "blk", "pf")
cor_matrix <- cor(nba[, predictors])
diag(cor_matrix) <- NA

cat("Strongest correlation between two predictors:",
    round(max(abs(cor_matrix), na.rm = TRUE), 3),
    "(assists and field goal %)\n")
```

Assists and field goal percentage correlate at
`r round(cor(nba$ast, nba$fg_pct), 3)` with each other — which is exactly
the mechanical relationship we'd expect, not a coincidence. This means the
marginal correlations above can overstate how much *independent* information
a stat like assists actually contributes once we already know a team's
shooting percentage. To account for this, we fit a multiple logistic
regression that estimates each stat's association with winning *while
holding the others constant*.

It's also worth being clear about what this analysis can and can't say. All
of these are box-score numbers recorded *during* the same game whose outcome
we're predicting — this is a description of what tends to co-occur with
winning within a game, not a forecast built from information known before
tip-off, and not a causal claim (a team that is already winning by a lot may
also rack up steals and fast-break assists *because* it's winning, not only
the reverse). Home court is included as a control, but we can't rule out
other confounders like opponent quality or pace of play that aren't in this
box-score-only dataset.

```{r logistic-regression}
nba_scaled <- nba %>%
  mutate(across(all_of(predictors), ~ as.numeric(scale(.)))) %>%
  mutate(home_num = as.integer(home))

fit <- glm(
  win ~ fg_pct + fg3_pct + oreb + dreb + ast + tov + stl + blk + pf + home_num,
  data = nba_scaled, family = binomial()
)

summary(fit)

mcfadden_r2 <- 1 - fit$deviance / fit$null.deviance
cat("McFadden pseudo R-squared:", round(mcfadden_r2, 3), "\n")
```

```{r regression-chart, fig.height=5}
coef_tbl <- tibble(
  stat = names(coef(fit))[-1],
  estimate = coef(fit)[-1]
) %>%
  arrange(desc(estimate))

ggplot(coef_tbl, aes(x = reorder(stat, estimate), y = estimate,
                      fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#1d428a", "FALSE" = "#c8102e"), guide = "none") +
  labs(title = "Standardized Logistic Regression Coefficients",
       subtitle = "Association with winning, holding other stats constant",
       x = NULL, y = "Coefficient (log-odds per 1 SD)") +
  theme_minimal()
```

This is where the skepticism pays off: **assists flips from one of the
strongest positive marginal correlations to a small *negative* coefficient**
once field goal percentage is held constant. Assists aren't bad for winning
— but almost all of the "assists help you win" signal turns out to already
be captured by field goal percentage; a team with a below-average number of
assists relative to its shooting percentage tends to do slightly worse, not
better. Field goal percentage and defensive rebounding remain the two
strongest positive predictors even after controlling for everything else,
and turnovers remain clearly negative. The model's McFadden pseudo
R-squared of about `r round(mcfadden_r2, 2)` indicates these box-score stats
jointly explain a substantial share of the win/loss outcome, but well short
of all of it — consistent with the idea that game outcomes also depend on
things box scores don't capture, like the opponent's stats in the same game.

# Conclusions

Returning to the original questions:

1. **Which team statistics have the strongest relationship with winning?**
   Field goal percentage and defensive rebounding, both on their own and
   after controlling for other stats. Turnovers and personal fouls are the
   only stats consistently *negatively* associated with winning.

2. **Are rebounds, assists, or shooting percentages more important?**
   Shooting efficiency (field goal % and three-point %) and defensive
   rebounding matter most. Assists look important in isolation, but that
   effect is largely explained by shooting percentage rather than
   independent playmaking value — a good example of why it matters to
   check correlated predictors rather than stopping at one-at-a-time
   correlations.

3. **Do winning teams commit fewer turnovers?**
   Yes. Winning teams average
   `r round(summary_tbl$TOV[summary_tbl$result == "Win"], 1)`
   turnovers per game versus
   `r round(summary_tbl$TOV[summary_tbl$result == "Loss"], 1)`
   for losing teams, and turnovers have a negative, statistically
   significant coefficient in the regression even after controlling for
   every other stat.

4. **Which statistics differ the most between winning and losing teams?**
   By both marginal correlation and regression coefficient, field goal
   percentage and defensive rebounding show the largest, most consistent
   gaps between winning and losing teams. Offensive rebounding is the
   biggest surprise: it looks irrelevant on its own but becomes a
   meaningfully positive predictor once shooting percentage is held
   constant — likely because offensive rebounds are mechanically more
   available after a missed shot, so their value is masked at the
   marginal level.

Overall, the data supports a fairly intuitive basketball story — shoot
efficiently, control the defensive glass, and take care of the ball — but
getting there required being skeptical of the raw, one-at-a-time
correlations, since several of these stats move together for reasons that
have nothing to do with winning.

# Reproducibility

This report reads `data/regular_season_totals_2010_2024.csv` from a path
relative to this file, so it will knit successfully as long as that CSV is
committed in the same repository under `data/`. It uses only CRAN packages
available for this course (`dplyr`, `tidyr`, `readr`, `stringr`, `ggplot2`,
`knitr`) and does not require network access at knit time.

```{r session-info}
sessionInfo()
```
