---
title: 'DS 2020 - Summer 2026 - Homework #3'
author: "David Lawlor"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab: Avengers' Peril

## Overview

In this activity we are going to

  1. reshape a data set, and then
  
  2. use the data to fact-check some statistics in a published report.
  
## Data Background

The data was collected by FiveThirtyEight. This time we are interested in a review of life and death among the Avengers, a group of super heroes in the Marvel universe. The original FiveThirtyEight article is no longer available through the original link. For this assignment, use the dataset and accompanying data description link below.

The description of the data is [here](https://github.com/fivethirtyeight/data/tree/master/avengers).

In a universe where time travel, alternate lifelines and portals with all kinds of powers exist, dying might be less permanent than in the world that we are used to.

In 2015 FiveThirtyEight has conducted a comprehensive review of all Avengers. We can access the resulting data using the command

```{r}
av <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/avengers/avengers.csv", stringsAsFactors = FALSE)
head(av)
```

## Data Tidying

The avenger data consists of ten columns for each avenger called Death[1-5] to Return[1-5]

Get the data into a format where the five columns for Death[1-5] are replaced by two columns: Time, and Death. Time should be a number between 1 and 5 (look into the function `parse_number()`); Death is a categorical variables with values "yes", "no" and "". Call the resulting data set `deaths`.

> Work here to reshape the data set.

```{r}
library(tidyverse)

av <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/avengers/avengers.csv", 
               stringsAsFactors = FALSE)

deaths <- av %>%
  pivot_longer(cols = Death1:Death5, 
               names_to = "Time", 
               values_to = "Death") %>%
  mutate(Time = parse_number(Time))

deaths
```

Similarly, deal with the returns of characters. Call the resulting data set `returns``

> Work here to reshape the data set.

```{r}
returns <- av %>%
  pivot_longer(cols = Return1:Return5,
               names_to = "Time",
               values_to = "Return") %>%
  mutate(Time = parse_number(Time))

returns
```

Answer the question: how many deaths on average, does an Avenger suffer?

> Answer here, and include all work and code done to get the answer.

```{r}
avg_deaths <- deaths %>%
  filter(Death == "YES") %>%
  group_by(URL) %>%
  summarise(num_deaths = n()) %>%
  summarise(avg = mean(num_deaths))

avg_deaths
```
What this does is first filter to only rows where an actual death happened, then counts deaths per Avenger, then takes the average across all Avengers.Doesnt count 0's

## Fact Checking

Pick one of the statements in the FiveThirtyEight [analysis](https://fivethirtyeight.com/features/avengers-death-comics-age-of-ultron/) and fact check it based on the data. Use `dplyr` functionality whenever possible.

> Answer here, and include all work and code done to get the answer.

```{r}
# total Avengers are in the dataset
nrow(av)

# how many have died at least once
deaths %>%
  filter(Death == "YES") %>%
  distinct(URL) %>%
  nrow()
```
The FiveThirtyEight article claimed there were 173 Avengers in the dataset and that 69 of them have died at least once. Both of these check out.

# Homework Assignment (Part 1)

## Flying etiquette

FiveThirtyEight is a website founded by Statistician and writer Nate Silver to publish results from  opinion poll analysis, politics, economics, and sports blogging. One of the featured articles considers [flying etiquette](https://fivethirtyeight.com/features/airplane-etiquette-recline-seat/).  This article is based on data collected by FiveThirtyEight and publicly available on github. Use the code below to read in the data from the survey:
```{r}
fly <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/flying-etiquette-survey/flying-etiquette.csv")
```

The following couple of lines of code provide a bit of cleanup of the demographic information by reordering the levels of the  corresponding factor variables. Run this code in your session.

```{r}
fly$Age <- factor(fly$Age, levels=c("18-29", "30-44", "45-60", "> 60", ""))
fly$Household.Income <- factor(fly$Household.Income, levels = c("$0 - $24,999","$25,000 - $49,999", "$50,000 - $99,999", "$100,000 - $149,999", "150000", ""))
fly$Education <- factor(fly$Education, levels = c("Less than high school degree", "High school degree", "Some college or Associate degree", "Bachelor degree",  "Graduate degree", ""))
```

1. Download the RMarkdown file with these homework instructions to use as a template for your work. Make sure to replace "Your Name" in the YAML with your name.

2. Some people do not travel often by plane. Provide a (visual) breakdown of travel frequency (use variable `How.often.do.you.travel.by.plane.`). Reorder the levels in the variable by travel frequency from least frequent travel to most frequent. Draw a barchart of travel frequency and comment on it.  Exclude all respondents who never fly from the remainder of the analysis. How many records does the data set have now?

> Answer here.

```{r}
fly <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/flying-etiquette-survey/flying-etiquette.csv")

fly$Age <- factor(fly$Age, levels=c("18-29", "30-44", "45-60", "> 60", ""))
fly$Household.Income <- factor(fly$Household.Income, levels = c("$0 - $24,999","$25,000 - $49,999", "$50,000 - $99,999", "$100,000 - $149,999", "150000", ""))
fly$Education <- factor(fly$Education, levels = c("Less than high school degree", "High school degree", "Some college or Associate degree", "Bachelor degree", "Graduate degree", ""))

fly$How.often.do.you.travel.by.plane. <- factor(
  fly$How.often.do.you.travel.by.plane.,
  levels = c("Never", "Once a year or less", "Once a month or less",
             "A few times per month", "A few times per week", "Every day")
)

ggplot(fly, aes(x = How.often.do.you.travel.by.plane.)) +
  geom_bar() +
  coord_flip()

fly <- fly %>%
  filter(How.often.do.you.travel.by.plane. != "Never")

nrow(fly)
```
The barchart shows that the most common travel frequency among respondents is "Once a year or less", followed by "Once a month or less". Very frequent flyersa make up a small portion of respondents. After reordering the levels from least to most frequent travel, the distribution is clearly right skewed, meaning most people in the survey fly relatively infrequently. After removing respondents who never fly, the dataset has 874 records.

3. In the demographic variables (`Education`, `Age`, and `Houshold.Income`), replace all occurrences of the empty string "" by a missing value `NA`. How many responses in each variable do not have any missing values? How many responses have no missing values in any of the three variables? (Hint: think of the function `is.na`)

> Answer here.

```{r}
fly$Education[fly$Education == ""] <- NA
fly$Age[fly$Age == ""] <- NA
fly$Household.Income[fly$Household.Income == ""] <- NA

sum(!is.na(fly$Education))
sum(!is.na(fly$Age))
sum(!is.na(fly$Household.Income))

sum(!is.na(fly$Education) & !is.na(fly$Age) & !is.na(fly$Household.Income))
```
After replacing all empty strings with NA in the three demographic variables, the number of non-missing responses are 837 for Education, 843 for Age, and 698 for Household Income. Household Income has the most missing values by far, which makes sense since people tend to be more reluctant to share their income. When looking at all three variables together, only 696 respondents have complete information with no missing values in any of the three variables.

4. Run the command below and interpret the output. What potential purpose can you see for the chart? What might be a problem with the chart? Find at least one purpose and one problem.
```{r, fig.width=8, fig.height=3, message=FALSE}
library(ggplot2)
fly$Education = with(fly, factor(Education, levels = rev(levels(Education))))

ggplot(data = fly, aes(x = 1)) + 
  geom_bar(aes(fill=Education), position="fill") + 
  coord_flip() +
  theme(legend.position="bottom") +
  scale_fill_brewer() + 
  xlab("Ratio") 
```

> Answer here.

The chart shows the education breakdown of survey respondents as a stacked proportional bar. One potential purpose is that it lets you quickly see the overall composition of the sample by education level in a single glance, which is useful for understanding if the survey sample is representative of the general population. A potential problem is that it is hard to compare the sizes of the middle categories since they dont all start from the same baseline.

5. Rename the variable `In.general..is.itrude.to.bring.a.baby.on.a.plane.` to `baby.on.plane.`. How many levels does the variable `baby.on.plane` have, and what are these levels? Rename the level labeled "" to "Not answered". Bring the levels of `baby.on.plane` in an order from least rude to most rude. Put the level "Not answered" last. Draw a barchart of variable `baby.on.plane`.  Interpret the result. (This question is very similar to question 2, but preps the data for the next question)

> Answer here.

```{r}
fly$baby.on.plane <- fly$In.general..is.itrude.to.bring.a.baby.on.a.plane.

levels(factor(fly$baby.on.plane))

fly$baby.on.plane[fly$baby.on.plane == ""] <- "Not answered"

fly$baby.on.plane <- factor(fly$baby.on.plane,
  levels = c("No, not at all rude", "Yes, somewhat rude", 
             "Yes, very rude", "Not answered")
)

# barchart
ggplot(fly, aes(x = baby.on.plane)) +
  geom_bar() +
  coord_flip()
```
There are 4 levels: "No, not at all rude", "Yes, somewhat rude", "Yes, very rude", and "Not answered". The empty string level was renamed to "Not answered" and placed last. The barchart shows that the overwhelming majority of respondents (just under 600) do not think bringing a baby on a plane is rude at all. A smaller group of around 180 find it somewhat rude, and only about 75 find it very rude. Very few people chose not to answer. This shows most people are tolerant of babies on planes.

6. Investigate the relationship between gender and the variables `Do.you.have.any.children.under.18.` and `baby.on.plane`. How is the attitude towards babies on planes shaped by gender and own children under 18? Find a plot that summarises your findings (use `ggplot2`).

> Answer here.

```{r}
ggplot(fly, aes(x = baby.on.plane, fill = Do.you.have.any.children.under.18.)) +
  geom_bar(position = "fill") +
  coord_flip() +
  facet_wrap(~Gender) +
  theme(legend.position = "bottom") +
  xlab("Proportion")
```
The plot shows a clear relationship between having children under 18 and attitudes toward babies on planes. Respondents who have children are more likely to say bringing a baby on a plane is not rude at all, with over 25% of that "not rude" group having kids. As attitudes become more negative toward babies on planes, the proportion of respondents with children drops, down to roughly 10-25% for somewhat rude and only 5-10% for very rude. This makes intuitive sense as parents are probably more sympathetic toward other parents traveling with babies. The pattern holds for both men and women, though women in the sample appear to have children slightly less often than men overall. Gender itself does not seem to dramatically change the relationship between having kids and attitudes toward babies on planes.

# Homework Assignment (Part 2)

## Liquor Sales in Iowa

This dataset contains the spirits purchase information of Iowa Class “E” liquor licensees by product and date of purchase from January 1, 2017 to current. The dataset can be used to analyze total spirits sales in Iowa of individual products at the store level.

For all of the questions use functionality from the  `tidyverse` whenever possible.

1. Download the RMarkdown file with these homework instructions to use as a template for your work. Make sure to replace "Your Name" in the YAML with your name.

2. The code below downloads the data from the Iowa Portal and saves a local copy to your machine. The first time you run this code, make sure you have a good internet connection and enough space on your machine (~50 MB). If the local copy exists, re-knitting the file will load the (binary) file from disk and be much faster. 

```{r, cache=TRUE}
if (!file.exists("ames-liquor.rds")) {
  url <- "https://github.com/ds202-at-ISU/materials/blob/master/03_tidyverse/data/ames-liquor.rds?raw=TRUE"
  download.file(url, "ames-liquor.rds", mode="wb")
}
data <- readRDS("ames-liquor.rds")
```

3. Quick fact overview:

  - how many observations are in  `data`?
  
> Answer here.

```{r}
if (!file.exists("ames-liquor.rds")) {
  url <- "https://github.com/ds202-at-ISU/materials/blob/master/03_tidyverse/data/ames-liquor.rds?raw=TRUE"
  download.file(url, "ames-liquor.rds", mode="wb")
}
data <- readRDS("ames-liquor.rds")

nrow(data)
```
661945 Different observations in the data

  - how many different cities are in the data? (Variable `City`, careful, trick question!)
  
> Answer here.

```{r}
n_distinct(data$City)

# the trick is probably capitalization differences, check with:
data %>%
  count(City) %>%
  arrange(City)
```

At first glance there appear to be 2 different cities in the data, however this is misleading. Both "AMES" and "Ames" refer to the same city. 

AMES    247979            
Ames    413966

  - different stores: how many different stores are in the data? Check first with `Store Name`, then with `Store Number`. Discuss differences (give an example), and then answer the question of how many different stores are in the data set.
  
> Answer here.

```{r}
# check by store name
n_distinct(data$`Store Name`)

# check by store number
n_distinct(data$`Store Number`)

# look at differences
data %>%
  count(`Store Name`, `Store Number`) %>%
  arrange(`Store Number`)
```

I counted the number of unique stores using both Store Name and Store Number and found that the results differ because some stores have inconsistent names. For example, "Hy-Vee #2 / Ames" and "Hy-vee #2 / Ames" share the same store number (2501), so Store Number provides a more accurate count of distinct stores in the dataset.

Note: Your submission is supposed to be fully reproducible, i.e. the TA and I will 'knit' your submission in RStudio.

For the submission: submit your solution in an R Markdown file and (just for insurance) submit the corresponding html (or Word) file with it.

(Optional but encouraged):  
If you’d like to practice using GitHub, feel free to push your `.Rmd` and knitted `.html` file to a **public GitHub repository** under your own account. If you do, paste the link to your GitHub repo below:

> GitHub repo link (optional): `__________`